---
title: 'Programming in R Exam Part 2'
subtitle: 'An Analysis of the Hitters Data'
author: "Emanuel Y. Gloria"
date: "December 27, 2024"
output:
  beamer_presentation:
    theme: "AnnArbor"
    colortheme: "dolphin"
    fonttheme: "structurebold"
    slide_level: 2
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

---

## Introduction

### Analysis of Baseball Data for the Year 1986

This presentation summarizes the analysis conducted on the 1986 baseball dataset, focusing on the relationship between `Walks` and `Runs`, and a statistical model that includes both variables along with the `Division` factor.

---

## Objective

The objective of this analysis is to understand the relationship between `Walks` and `Runs` in the context of the 1986 baseball data. The analysis is also extended by including the `Division` factor (East and West Divisions) to assess how these factors influence the number of runs scored.

---

## Results

### Scatter Plot

Plotting the relationship between `Walks` and `Runs` and extending it by including the `Division` factor (East and West Divisions). This shows a positive linear trend.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Load necessary libraries
library(ggplot2)
library(ISLR)
library(dplyr)

data(Hitters)

# Create a scatterplot between Walks and Runs, disaggregated by Division
scatterplot <- ggplot(Hitters, aes(x = Walks, y = Runs, color = Division)) +
  geom_point(size = 1.75, alpha = 0.8) +  # Add points with specified size
  scale_color_manual(values = c("E" = "red", "W" = "black")) +  # Color E points red, W points black
  labs(
    x = "Walks",
    y = "Runs",
    color = "Division"  # Label for the legend
  ) +
  theme_minimal() +
  theme(
    plot.margin = margin(t = 20, r = 20, b = 50, l = 20),  # Adjust bottom margin to create space for subtitle
    plot.subtitle = element_text(hjust = 0, vjust = 1, size = 10, face = "plain"),  # Left-align the subtitle
    plot.caption = element_text(hjust = 0, vjust = 0, size = 10, face = "plain")  # Add custom caption as bottom-left label
  )

# Add the label "Figure 1.2" using caption (treated as subtitle outside the plot area)
scatterplot <- scatterplot +
  labs(caption = "Figure 1.2")  # This moves the label to the bottom-left corner

# Show the plot
scatterplot
```
---
## Results

### Correlation Analysis

Calculating the correlation between Walks and Runs by Division and rounded it to three decimal places revealed a score of at least 0.7 indicating a strong positive relationship.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Calculate the correlation between Walks and Runs by Division and round the correlation to 3 decimal places
correlation_by_division <- Hitters %>%
  group_by(Division) %>%
  summarize(Correlation = round(cor(Walks, Runs, method = "pearson"), 3)) %>%
  as_tibble()  # Convert to tibble for better display

# Print the result
print(correlation_by_division)
```